---
title: "Google Capstone Project Cyclist data"
author: "Samay Jain"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Importing the data sets
```{r, message = FALSE}
setwd("C:/Users/HP/Downloads/Capstone project/cyclist data")
Apr_22 = read_csv("202204-divvy-tripdata.csv")
May_22 = read_csv("202205-divvy-tripdata.csv")
Jun_22 = read_csv("202206-divvy-tripdata.csv")
Jul_22 = read_csv("202207-divvy-tripdata.csv")
Aug_22 = read_csv("202208-divvy-tripdata.csv")
Sep_22 = read_csv("202209-divvy-publictripdata.csv")
Oct_22 = read_csv("202210-divvy-tripdata.csv")
Nov_22 = read_csv("202211-divvy-tripdata.csv")
Dec_22 = read_csv("202212-divvy-tripdata.csv")
Jan_23 = read_csv("202301-divvy-tripdata.csv")
Feb_23 = read_csv("202302-divvy-tripdata.csv")
Mar_23 = read_csv("202303-divvy-tripdata.csv")
```

# Combining the data frame
```{r, message = FALSE}
Merged_data <- rbind(Apr_22, May_22, Jun_22, Jul_22, Aug_22, Sep_22, Oct_22, Nov_22, Dec_22, Jan_23, Feb_23, Mar_23 )
```

# Viewing The data set
```{r}
head(Merged_data)
```

# Renaming the Columns
```{r}
colnames(Merged_data)[1] <- "Ride_ID"
colnames(Merged_data)[2] <- "Bike_type"
colnames(Merged_data)[3] <- "Start_time"
colnames(Merged_data)[4] <- "End_time"
colnames(Merged_data)[13] <- "Member_type"
```

# Mutating the columns
```{r}
Merged_data$Ride_time <- difftime(Merged_data$End_time, Merged_data$Start_time, units = "mins")
Merged_data$Day <- weekdays(Merged_data$Start_time)
Merged_data$Month <- format(as.Date(Merged_data$Start_time), "%m")
Merged_data$Hour <- format(Merged_data$Start_time, "%H")
Merged_data$Distance <- (sqrt((Merged_data$end_lat - Merged_data$start_lat)**2 + (Merged_data$end_lng - Merged_data$start_lng)**2))*111
```


## Dropping Extra columns
```{r}
Merged_data <- Merged_data[, -c(6, 8)]
```

# Structure of data
```{r}
str(Merged_data)
```

# Performing data type conversion
```{r}
Merged_data$Bike_type <- factor(Merged_data$Bike_type, levels = c("electric_bike", "classic_bike", "docked_bike"), 
                                 labels = c("Electric", "Classic", "Docked"))
Merged_data$Member_type <- factor(Merged_data$Member_type, levels = c("member", "casual"), labels = c("Member", "Casual"))
#Merged_data$Month <- factor(Merged_data$Month, levels = c("01", "02", "03", "04", "05", "06", "07", "08", "09", "10", "11", "12"), labels = c("Jan", #"Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))
```


# Scanning data for inconsistencies
```{r}
summary(as.numeric(Merged_data$Ride_time))
sd(Merged_data$Ride_time)
summary(as.numeric(Merged_data$Distance))
sd(Merged_data$Distance, na.rm = TRUE)

```

# Calculating outliers range for variables (3rd quartile - 1st quartile +- 1.25(std dev.))
```{r}
outlier_Ride_time <- 18.02 - 5.68 + 1.25*173.7513
outlier_Distance <- 3.140 - 1.029 + 1.25*12.83396
outlier_Ride_time
outlier_Distance 
```

# Removing outliers and inconsistencies(if any) from the data set
```{r}
Merged_data <- Merged_data[Merged_data$Ride_time > 0 & Merged_data$Ride_time < 230, ]
Merged_data <- Merged_data[Merged_data$Distance > 0 & Merged_data$Distance < 18.5, ]
Merged_data <- na.omit(Merged_data)
```

# Checking for any inconsistencies
```{r}
summary(as.numeric(Merged_data$Ride_time))
sd(Merged_data$Ride_time)
summary(as.numeric(Merged_data$Distance))
sd(Merged_data$Distance, na.rm = TRUE)
```

# Saving the data file
#```{r}
#write.csv(Merged_data, file = "Merged_Data.csv", row.names = FALSE)
#```

```{r}
#write.xlsx(Merged_data, file = "Merged_Data1.xlsx", rowNames = FALSE)
```

# Visualization
#```{r}
#Member_character <- as.character(Merged_data$member_casual)
#Member_type <- table(Member_character)
#barplot(Member_type)
#```
#```{r}
#Merged_data %>% ggplot(Member_character, aes(x = Day, fill = Member_character)) +
#  geom_bar(stat = "identity", position = "dodge") +
#  labs(title = "Number of Riders on Each Day", x = "Day") +
#  scale_fill_manual(values = c("blue", "red"), labels = c("Member", "Casual"))
#```
